FEM Transition Models

### 2023-03-11 FEM_R/Loneliness_Social_Isolation/transition_models_testing.Rmd
Massive improvements to the loneliness model including lots of new variables, as well as changes to the reference factors for lnly and sociso which are particularly important in the mortality model. IT WORKS

require(tidyverse)
require(haven)

@@

# make sure mortality if factor
trans$died <- factor(trans$died)

@@

sociso.check <- sociso %>% 
  group_by(l2sociso) %>%
  mutate(total = sum(Freq)) %>%
  mutate(prop = (Freq / total) * 100)
```

@@

@@

```{r}
form.mort.min <- 'died ~ male + white + hsless + college + 
                  l2age65l + l2age6574 + l2age75p + 
                  l2lnly2 + l2lnly3 +
                  l2sociso2 + l2sociso3 + l2sociso4 + l2sociso5 + l2sociso6'

>>>>

require(tidyverse)
require(haven)
require(MASS)

@@

# make sure responses are factor where necessary
trans$died <- factor(trans$died)
trans$lnly <- factor(trans$lnly)
trans$sociso <- factor(trans$sociso)
trans$male <- factor(trans$male)
trans$hsless <- factor(trans$hsless)
trans$college <- factor(trans$college)
trans$l2employed <- factor(trans$l2employed)
trans$l2inactive <- factor(trans$l2inactive)
trans$l2physact <- factor(trans$l2physact)

@@

sociso.check <- sociso %>% 
  group_by(l2sociso) %>%
  mutate(total = sum(Freq)) %>%
  mutate(prop = (Freq / total) * 100)

#write.csv(lnly.check, )
```

@@

Lets have one more go with these limited to waves 1-6, as no more deaths are recorded after wave 6.

```{r}
trans2 <- trans %>% filter(wave < 7)

lnly2 <- as.data.frame(xtabs(formula = ~died + l2lnly, data = trans2, addNA = TRUE))
sociso2 <- as.data.frame(xtabs(formula = ~died + l2sociso, data = trans2, addNA = TRUE))

lnly.check2 <- lnly2 %>%
  group_by(l2lnly) %>%
  mutate(total = sum(Freq)) %>%
  mutate(prop = (Freq / total) * 100)

sociso.check2 <- sociso2 %>%
  group_by(l2sociso) %>%
  mutate(total = sum(Freq)) %>%
  mutate(prop = (Freq / total) * 100)
```

@@

```{r}
form.mort.ls <- 'died ~ male + white + hsless + college + 
                  l2age65l + l2age6574 + l2age75p + 
                  l2lnly1 + l2lnly2 + l2lnly3 +
                  l2sociso1 + l2sociso2 + l2sociso3 + l2sociso4 + l2sociso5 + l2sociso6'

probit.mort.ls <- glm(formula = form.mort.ls, family = binomial(link = 'probit'),
                    data = trans)

summary(probit.mort.ls)
```

Interesting. Mirrors the FEM transition model, although not exactly. Same sign for each var and similar relative magnitude (i.e. in the sociso levels). The sociso relationship is great now but the loneliness is not so clear cut. This actually mirrors the literature so we could be on to something here! Try a version where the reference factors are different, might be good to pick the worst of each var and then we can see where the relationship falls down.

```{r}
form.mort.ls2 <- 'died ~ male + white + hsless + college + 
                  l2age65l + l2age6574 + l2age75p + 
                  l2lnly1 + l2lnly2 +
                  l2sociso1 + l2sociso2 + l2sociso3 + l2sociso4 + l2sociso5'

probit.mort.ls2 <- glm(formula = form.mort.ls2, family = binomial(link = 'probit'),
                    data = trans)

summary(probit.mort.ls2)
```

*Chefs kiss*. Exaclty what we want to see. Probability of mortality increases with higher rates of loneliness and social isolation. Appears that loneliness here is more highly associated with mortality than social isolation, but I have an idea about that. Perhaps, with loneliness being the subjective feeling and social isolation the objective situation, social isolation is more likely in people in palliative care or close to end of life already. These people are harder to follow in ELSA, and institutionalised people are removed from the survey. Maybe a contributing factor.

Try one more version, with chronic diseases and stuff.

```{r}
form.mort.full <- 'died ~ male + white + hsless + college + 
                  l2age65l + l2age6574 + l2age75p + 
                  l2lnly1 + l2lnly2 +
                  l2sociso1 + l2sociso2 + l2sociso3 + l2sociso4 + l2sociso5 +
                  l2cancre + l2diabe + l2hearte + l2alzhe + l2demene + l2lunge +
                  l2stroke'

probit.mort.full <- glm(formula = form.mort.full, family = binomial(link = 'probit'),
                    data = trans)

summary(probit.mort.full)
```

# Loneliness and Social Isolation

Now lets have a go at fitting models to lnly and sociso.

## Loneliness

```{r}
# Get a sample comparable to the sample used in FEM
trans.lnly <- trans %>%
  dplyr::select(idauniq, wave, died, lnly, male, white, hsless, college, l2age65l, l2age6574, l2age75p, atotb, l2employed, l2inactive, l2physact) %>%
  filter(wave > 1) %>%
  filter(died == 0)
trans.lnly.complete <- trans.lnly[complete.cases(trans.lnly),]

## MINIMAL
form.lnly.min <- 'lnly ~ male + scale(l2age65l) + scale(l2age6574) + scale(l2age75p)'
probit.lnly.min <- polr(formula = form.lnly.min,
                    data = trans.lnly,
                    Hess = TRUE,
                    na.action = na.omit)
print('Minimal model:')
summary(probit.lnly.min)

## FULL
form.lnly <- 'lnly ~ male + white + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + scale(atotb) + l2employed + l2inactive + l2physact'
probit.lnly <- polr(formula = form.lnly,
                    data = trans.lnly,
                    Hess = TRUE,
                    na.action = na.omit)
print('Full model:')
summary(probit.lnly)
```

Now use these models as a testbed using ideas from literature.

First up [Cohen-Mansfield et al (2015)](https://www.cambridge.org/core/journals/international-psychogeriatrics/article/abs/correlates-and-predictors-of-loneliness-in-olderadults-a-review-of-quantitative-results-informed-by-qualitative-insights/EF57448627F1EF5CFD89F83C3F188175).

Shortlist of variables I already have or can easily test:
- [x] gender
- [x] age
- [x] unmarried status
- [x] income (wealth?)
- [x] education
- nursing home resident (may be issues with 0 analysis weight) (nhmliv)
- [x] participates in social activities (socyr)
- [x] care for someone in household (R*GCAREINHH1W)
- [x] Childless (child)
- [x] living alone / few individuals in household (H*HHRES)
- [x] self reported health == poor (SRH == 1)
- [x] functional status == poor (ADL == 3+, IADL == 2+)
- [x] depression (psychiatric problems - emotional, nervous, or psychiatric)

```{r}
## TEST
form.lnly.test <- 'lnly ~ male + hsless + college + scale(l2age65l) + scale(l2age6574) + scale(l2age75p) + l2cohab + l2widowed + l2single + l2employed + l2inactive + l2physact + l2anyadl + l2anyiadl + l2srh5 + l2psyche + l2hhres + l2socyr + l2gcareinhh1w + childless + scale(l2logatotb)'
probit.lnly.test <- polr(formula = form.lnly.test,
                    data = trans,
                    Hess = TRUE,
                    na.action = na.omit)
print('Test model:')
summary(probit.lnly.test)
```

```{r}
